The brightness and spatial distributions of terrestrial radio sources

### 2012-12-19 articles/Article-RFI-distribution/article-rfi-distribution.tex
Updating article with Wim Brouw's comments

\ifpdf\pdfinfo{/Title      (The brightness and spatial distributions of terrestrial radio sources)
               /Author     (A. R. Offringa et al.)
               /Keywords   (dark ages, reionization, first stars;atmospheric effects;instrumentation: interferometers;methods: observational;techniques: interferometric;radio continuum: general;rfi)
        }
\else\usepackage{graphics}\fi

@@

\begin{abstract}
Feeble undetected sources of radio-frequency interference (RFI) might become visible after integration of long radio observations if they are coherently present over time. Thereby, they might obstruct the detection of the astronomical signal of interest. This issue is especially important for Epoch of Reionization (EoR) projects that try to detect the faint redshifted HI signals of the earliest objects in the Universe. We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR. The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.6 to -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present. Since these observations have achieved the thermal (sky) noise limit, a logical implication is that most of the RFI sources do not add coherently.
\end{abstract}

\begin{keywords}
atmospheric effects -- instrumentation: interferometers -- methods: observational -- techniques: interferometric -- radio continuum: general -- dark ages, reionization, first stars
\end{keywords}

@@

Typical radio observations record a few hours of data, and the results are integrated. In these cases, it often suffices to only excise interference that is apparent and thus above the noise to reach the thermal noise limit of the instrument. A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated. In such a case, feeble interference caused by stationary sources might not manifest itself above the noise in individual observations, but is coherently present in the data. Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest. This is very relevant for the 21-cm Epoch of Reionisation (EoR) experiments, because they involve long integration times. Several such experiments are underway, to either measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or the global signal \citep{edges}. For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest. In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array \citep{de-bruyn-eor-ursi-2011}.

@@

\section{Predicting the brightness distribution} \label{sec:distribution-prediction}
Interference is generated by many different kind of transmitters, and these will have different spatial and brightness distributions (``spatial'' refers here to the distribution on the Earth). For example, aeroplanes and satellites have widely different heights, while other sources are ground-based. Even ground-based sources might be spread differently. For example, it can be expected that citizens' band (CB) devices, that are often used in cars, are differently distributed from broadcasting transmitters. The frequencies of deliberately transmitting sources of a particular class of devices are constrained by the bands that have been allocated for the given class, which might allow one to distinguish different sources to some extent.

@@

In every dynamic spectrum we can measure the number of times that the flux density is within a particular range. Dividing this quantity by the total number of samples yields the relative number of events as a function of intensity. We will refer to this quantity with the statistical term ``frequency density'' (not to be confused with the physical radiation frequency). We will now start by deriving a prediction of the frequency density function of ground-based interfering sources. Consider an interfering point source of strength $I$. This source is observed by an interferometer that consists of two antennae or stations with gains $g_1,g_2$, which include all instrumental effects. The antennae are located at distances $r_1,r_2$ from the source. The interferometer will record an apparent instantaneous strength $\mathcal{S}$ of
\begin{equation}
 \mathcal{S}(r_1, r_2) = I \frac{g_1 g_2}{r_1 r_2},
\end{equation}
with $g_1,g_2>0$ and $r_\textrm{L} > r_1,r_2 > 0$. Here, $r_\textrm{L}$ is a limiting distance, which will be well below the diameter of the Earth. We will limit our analysis to cross-correlated antennae; the auto-correlations will be ignored.

@@

Now, we can treat the interferometer geometrically as a single point, as both antennae will see the same distribution. Then, we can express the received amplitude $\mathcal{S}$ for a given distance $r$ and gain $g$ as
\begin{equation} \label{eq:amplitude-fall-off-in-free-space}
 S(r) = \frac{Ig}{r^2}.
\end{equation}

Next, we assume that all sources have equal constant strength $I$ and follow a uniform spatial distribution in a two-dimensional plane. This is obviously a simplification, as the sources are actually distributed on the surface of a sphere. Hence, beyond some distance, the Earth will partly block the path between transmitter and receiver. Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid. Using these assumptions, we can express the expected cumulative frequency density of sources $f_I(r)$ at distance $r$ as
\begin{equation} \label{eq:cumulative-distance-function}
F_{\textrm{distance} \le r}(r)=c\pi r^2,
\end{equation}
for some constant $c$ that represents the number of sources per unit area. In other words, we will observe $F$ sources that are at most a distance of $r$ away.

@@

\subsection{Spherical case}
If we do take into account the fact that the sources lie on a sphere instead of a plane, Eq.~\eqref{eq:cumulative-distance-function} changes. In this case, the number of sources corresponds with the surface area of a hemisphere. By using the formula for the surface area of a hemisphere and some basic geometry, one finds that
\begin{equation}\label{eq:cumulative-spherical}
\tilde{F}_{\textrm{distance} \le r} = c 2\pi R^2\left(1-\cos(2\arcsin \frac{r}{2R})\right) \hspace{0.4cm}\textrm{with } r \le 2R,
\end{equation}
with $R$ the radius of the Earth. Eqs.~\eqref{eq:cumulative-distance-function} and \eqref{eq:cumulative-spherical} have the same value over the range they can be evaluated. Therefore, the spherical and 2-dimensional cases lead to a distribution with equal power law. Evaluating Eq.~\eqref{eq:cumulative-spherical} is limited to $r \le 2R$, but in practice the lowest amplitude of observed sources is further constrained because of the curvature of the Earth. This causes distant sources to settle below the horizon. Nevertheless, some of those sources might still be visible due to reflection of the ionosphere and other (semi-)spherical propagation effects.

\subsection{Propagation effects}
So far, we have assumed that the electromagnetic radiation propagates through free space, resulting in a $r^{-2}$ fall-off. In reality, the radiation will be affected by complicated propagational effects. Because of the irregular surface of the Earth (including urban areas) and the absence of a direct line of sight between transmitter and receiver, the propagation mode might be indirect. Multiple indirect paths might be formed by reflection, refraction or diffraction of the electromagnetic wave.

A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}. The original model is based on urban areas, but corrections are given by \citeauthor{hata-propagation-loss} for sub-urban areas with a lower population density and for open areas. \citeauthor{hata-propagation-loss} gives the following analytical estimate for $L_p$, the electromagnetic propagation loss over land:
\begin{align} 
\notag                L_p & = & 69.55 + 26.16 \log_{10} f_c - 13.82 \log_{10} h_b - \\
\label{eq:hata-model} & &  a(h_m) + (44.9 - 6.55 \log_{10} h_b) \log_{10} r, 
\end{align}
where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in meters; $h_m$ the height of the receiving antenna in meters; $r$ the distance between the antennae in meters; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the urban density. \citeauthor{hata-propagation-loss} found this model to be representative for frequencies $f_c \sim$~150--1500~MHz, with transmitter heights $h_b \sim$~30--200~m, receiver heights $h_m \sim$~1--10~m and over distances $r\sim$~1--20~km.

@@

So far, these are the expected histograms for pure noise and pure RFI that propagates through free space. However, the measured distribution is a mixture of the two. A perfect RFI detector would separate the samples in two distributions; one that is not affected by RFI, and therefore contains noise only, and one that is the sum of RFI and noise. Real RFI detectors can separate these distributions to some extent, but due to false positives and false negatives, the histograms will get mixed nevertheless. Even more, we still need to take into account that the RFI classified samples are also affected by noise. Although this effect is moderate, because 'most' RFI samples are of much higher amplitude than what is added because of the noise, it is the low-level RFI samples which we are interested in, and the relative effect of noise on those samples is large.

@@

To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as algebraic calculation is not trivial. The histogram can be numerically estimated by drawing complex real and imaginary samples from the two distributions and calculating and counting the amplitudes. A sample can be drawn from the RFI distribution by normalizing and inverting the cumulative frequency function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable. Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity. Therefore, to draw a sample by using the inversion method, this would require a uniformly distributed variable $\sim U(0,\infty)$. By assuming that there are no samples beyond some limiting distance $r_L$, then $F_{\textrm{amplitude}\le S}(S) = N-c \pi r_L^2$ for $S < Ig/r_L^2$. Consequently, by sampling $x_u$ from a uniform distribution with $0 \le x_u \le N-c \pi r_L^2$ and applying the inverse of Eq.~\eqref{eq:cumulative-rfi-distribution}, we will sample amplitudes $S$ with $S \ge Ig/r_L^2$, and $S\sim$ the power-law distribution.

By normalizing $x_u$, a single complex RFI contaminated sample $S_{\textrm{RFI}}$ can be sampled with
\begin{equation} \label{eq:drawing-rfi-distribution}
\textrm{real}(S_{\textrm{RFI}})\leftarrow \frac{Ig}{x_u r_L^2}\textrm{,\hspace{1mm} imag}(S_{\textrm{RFI}}) \leftarrow \frac{Ig}{y_u r_L^2},
\end{equation}
with $x_u, y_u \sim U(0, 1)$, and the corresponding formula for drawing a sample $S$ that is contaminated by both RFI and noise is
\begin{equation} \label{eq:drawing-combined-distribution}
\textrm{real}(S) \leftarrow x_n + \frac{Ig}{x_u r_L^2}
\textrm{,\hspace{1mm} imag}(S) \leftarrow y_n + \frac{Ig}{y_u r_L^2},
\end{equation}
with $x_n \sim N(\mu=0; \sigma=1)$. Of course, the final amplitude sample $S_A$ can be calculated with 
\begin{equation} \label{eq:drawing-amplitudes}
S_A \leftarrow |S| = \sqrt{\textrm{real}({S})^2+\textrm{imag}({S})^2}.
\end{equation}
An example of distribution curves for various settings of $Ig/r_L^2$ is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.

@@

One instrumental effect that is absorbed in $g$ is the frequency response of the instrument, i.e., the antenna response in combination with the passband of the analogue and digital filters. Because the data that are analysed in Sect.~\ref{sec:dist-results} have initially not been band-pass calibrated, the instrumental response is not uniform over frequency. We determined that the variation due to the band-pass is about one order of magnitude in the LBA and about a factor of two in the HBA.

@@

%In Fig.~\ref{fig:plot-passband-histogram}, the histograms of the frequency response are given. To also demonstrate the effect of mixing distributions, histograms are added for samples that are drawn from a uniform, Gaussian and -2 power-law distribution as in Eq.~\eqref{eq:drawing-rfi-distribution} with $Ig/r_L^2=1$, that have been multiplied by the passband. The frequency responses were determined by measuring the standard deviation of the data over frequency after flagging the data for interference. As expected, it can still be determined that the intrinsic distribution is a -2 power law.

Another effect that is absorbed in $g$, is the beam of the instrument. At the point of writing, LOFAR beam models are still being developed and are not yet well parametrized. However, since it is likely that most RFI sources are observed at the edges of the beam, it can be expected that the beam will have a limited effect on the histogram of an observation, comparable with the effect of the frequency response.

@@

\subsection{Estimating $\sigma$ and slope parameters}
The mode $\hat\sigma$ of the Rayleigh distribution is estimated by finding the amplitude with the maximum occurrences, i.e., the amplitude corresponding to the peak of the histogram. The slope is estimated using linear regression over a visually selected interval.
%\begin{equation}
%R=\left[20\hat\sigma; 10^{\frac{1}{2}\left(\log_{10} 20\hat\sigma + \log_{10} S_\textrm{max}\right) } \right].
%\end{equation}
%In terms of the log-log histogram, the start of the interval corresponds with the point that is $1.3$ units to the right of the peak in the histogram, and the interval end corresponds with the point halfway between the start point and the maximum amplitude. This interval starts near the start of the RFI slope but past where the Rayleigh curve dominates. It also does not include the noisy data at the end of the curve. The curves are not always completely straight in this interval, but since we use the same stable way to determine the interval over which to calculate the slope, we can at least compare them.

One should note that fitting straight lines to the distribution curve in a log-log plot is not the most accurate way of estimating the exponent of a power-law distribution \citep{power-law-distribution}. However, because of our enormous sample size, which allows fitting the line over a large interval, the estimator will be sufficiently accurate for our purpose. Nevertheless, we will additionally calculate a maximum likelihood estimator for comparison. The maximum-likelihood estimator for the exponent in a power-law distribution is given by the Hill estimator \citep{hill-estimator, power-law-distribution}, defined as:
\begin{equation} \label{eq:hill}
 \alpha_H = 1 + N \left(\sum\limits_{i=1}^{N} \ln \frac{x_i}{x_\textrm{max}} \right)^{-1}.
\end{equation}
However, this estimator assumes the distribution is not cut-off at a high point. In our case, the distribution is cut-off at both sides, for example because of the limit of the analogue-digital converter (see \S\ref{sec:rfi-distribution-constraints}). Therefore, using this estimator will result in an estimate that is steeper (i.e., more negative) than the actual distribution. Moreover, the estimator requires an iteration over all data points, and because of the size of the data sets this is somewhat impractical. However, we can calculate Eq.~\eqref{eq:hill} by adding the bin centre values $N$ times to the calculation, where $N$ the bin frequency. Because our bin size is small and the distribution is well sampled, the resulting estimate will be close to the normal Hill estimator. This overcomes the need to iterate over all data.

@@

Similar to the calculation of $S_U$ and by using $N_\textrm{total}$ as an upper limit to the total number of samples that are affected by the power-law distribution, one can estimate the lower limit $S_L$. For Fig.~\ref{fig:explanation-constraints} this means that the areas labelled $N_a$ and $N_\textrm{total}-N_\textrm{part}$ are equal. Solving this equality, one finds that
\begin{equation} \label{eq:lower-limit-1}
S_L = \sqrt[\alpha+1]{\frac{\alpha+1}{\beta}\left(N_\textrm{part} - N_\textrm{total} \right) + S_1^{\alpha+1}}.
\end{equation}

@@

\section{Data description} \label{sec:dist-data}
We have analysed the distributions of two data sets. Both data sets are 24 h LOFAR RFI surveys and are extensively analysed in
\citet{lofar-radio-environment}. % TODO
In one set, the low-band antennae (LBA) were used and the frequency range 30.1--77.5 MHz was recorded, while in the other the high-band antennae (HBA) were used to record the frequency range 115.0--163.3 MHz. More stations were used in the LBA set. The specifications of the two sets are listed in Table~\ref{table:dist-data-specs}. The stations that have been used are geometrically spread over an area of about 80~km and 30~km in diameter at maximum for the LBA and HBA sets respectively.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=12cm]{img/histograms-raw-trimmed}
\caption{The histograms of the two data sets before pass-band correction and flux calibration.}
\label{fig:Histograms-raw}
\end{center}
\end{figure*}

@@

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/plot-lba-dist-per-frequency-trimmed}
\caption{Histograms for $5$ different $0.2$ MHz LBA sub-bands without pass-band correction and flux calibration. The continuous lines represent the data before RFI flagging. The dashes lines are the histograms of the samples that have been classified as RFI.}
\label{fig:plot-dist-per-frequency-LBA}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/plot-hba-dist-per-frequency-trimmed}
\caption{Histograms for $5$ different $0.2$ MHz HBA sub-bands without pass-band correction and flux calibration. The continuous lines represent the data before RFI flagging. The dashes lines are the histograms of the samples that have been classified as RFI.}
\label{fig:plot-dist-per-frequency-HBA}
\end{center}
\end{figure}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=13cm]{img/histograms-corrected-trimmed}
\caption{Observed LBA distribution after pass-band correction and flux calibration. $S_{l1}$ and $S_{l2}$ denote the limits of the distribution with a sharp lower cut-off (Eq.~\eqref{eq:lower-limit-1}) and uniform lower limit (Eq.~\eqref{eq:lower-limit-2}), $S_d$ is the average lower limit of RFI that is detected.}
\label{fig:histogram-passband-corrected}
\end{center}
\end{figure*}

>>>>

\ifpdf\pdfinfo{/Title      (The brightness and spatial distributions of terrestrial radio sources)
               /Author     (A. R. Offringa et al.)
               /Keywords   (dark ages, reionisation, first stars;atmospheric effects;instrumentation: interferometers;methods: observational;techniques: interferometric;radio continuum: general;rfi)
        }
\else\usepackage{graphics}\fi

@@

\begin{abstract}
Feeble undetected sources of radio-frequency interference (RFI) might become visible after integration of long radio observations if they are coherently present over time. Thereby, they might obstruct the detection of the astronomical signal of interest. This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals of the earliest objects in the Universe. We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR. The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.6 to -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present. Since these observations have achieved the thermal (sky) noise limit, a logical implication is that most of the RFI sources do not add coherently.
\end{abstract}

\begin{keywords}
atmospheric effects -- instrumentation: interferometers -- methods: observational -- techniques: interferometric -- radio continuum: general -- dark ages, reionisation, first stars
\end{keywords}

@@

Typical radio observations record a few hours of data, and the results are integrated. In these cases, it often suffices to only excise interference that is apparent and thus above the noise to reach the thermal noise limit of the instrument. A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated. In such a case, feeble interference caused by stationary sources might not manifest itself above the noise in individual observations, but might be coherently present in the data. Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest. This is very relevant for the 21-cm Epoch of Reionisation (EoR) experiments, because they involve long integration times. Several such experiments are underway, to either measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or the global signal \citep{edges}. For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest. In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array \citep{de-bruyn-eor-ursi-2011}.

@@

\section{Predicting the brightness distribution} \label{sec:distribution-prediction}
Interference is generated by many different kind of transmitters, and these will have different spatial and brightness distributions (``spatial'' refers here to the distribution on the Earth). For example, aeroplanes and satellites have widely different heights, while other sources are ground-based. Even ground-based sources might be spread differently. For example, it can be expected that citizens' band (CB) devices, that are often used in cars, are distributed differently from broadcasting transmitters. The frequencies of deliberately transmitting sources of a particular class of devices are constrained by the bands that have been allocated for the given class, which might allow one to distinguish different sources to some extent.

@@

In every dynamic spectrum we can measure the number of times that the flux density is within a particular range. Dividing this quantity by the total number of samples yields the relative number of events as a function of intensity. We will refer to this quantity with the statistical term ``frequency density'' (not to be confused with the physical radiation frequency). We will now start by deriving a prediction of the frequency density function of ground-based interfering sources. Consider an interfering point source of strength $I$. This source is observed by an interferometer that consists of two antennas or stations with gains $g_1,g_2$, which include all instrumental effects. The antennas are located at distances $r_1,r_2$ from the source. The interferometer will record an apparent instantaneous strength $\mathcal{S}$ of
\begin{equation}
 \mathcal{S}(r_1, r_2) = I \frac{g_1 g_2}{r_1 r_2},
\end{equation}
with $g_1,g_2>0$ and $r_\textrm{L} > r_1,r_2 > 0$. Here, $r_\textrm{L}$ is a limiting distance, which will be well below the diameter of the Earth. We will limit our analysis to cross-correlated antennas; the auto-correlations will be ignored.

@@

Now, we can treat the interferometer geometrically as a single point, as both antennas will see the same distribution. Then, we can express the received amplitude $\mathcal{S}$ for a given distance $r$ and gain $g$ as
\begin{equation} \label{eq:amplitude-fall-off-in-free-space}
 S(r) = \frac{Ig}{r^2}.
\end{equation}

Next, we assume that all RFI sources have equal constant strength $I$ and follow a uniform spatial distribution in a two-dimensional plane. This is obviously a simplification, as the RFI sources are actually distributed on the surface of a sphere with different heights. Beyond some distance, the Earth will partly block the path between transmitter and receiver. Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid. Using these assumptions, we can express the expected cumulative frequency density of sources $f_I(r)$ at distance $r$ as
\begin{equation} \label{eq:cumulative-distance-function}
F_{\textrm{distance} \le r}(r)=c\pi r^2,
\end{equation}
for some constant $c$ that represents the number of sources per unit area. In other words, we will observe $F$ sources that are at most a distance $r$ away.

@@

\subsection{Spherical case}
If we do take into account the fact that the sources lie on a sphere instead of a plane, Eq.~\eqref{eq:cumulative-distance-function} changes. In this case, the RFI sources are distributed over the surface of a hemisphere. If we assume their difference in height is negligible, and by using the formula for the surface area of a hemisphere and some basic geometry, one finds that
\begin{equation}\label{eq:cumulative-spherical}
\tilde{F}_{\textrm{distance} \le r} = c 2\pi R^2\left(1-\cos(2\arcsin \frac{r}{2R})\right) \hspace{0.4cm}\textrm{with } r \le 2R,
\end{equation}
with $R$ the radius of the Earth. Eqs.~\eqref{eq:cumulative-distance-function} and \eqref{eq:cumulative-spherical} have the same value over the range they can be evaluated. Therefore, the spherical and 2-dimensional cases lead to a distribution with equal power law. Evaluating Eq.~\eqref{eq:cumulative-spherical} is limited to $r \le 2R$, but in practice the lowest amplitude of observed sources is further constrained because of the curvature of the Earth. This causes distant sources to settle below the horizon. Nevertheless, some of those sources might still be visible due to reflection of the ionosphere and other (semi-)spherical propagation effects.

\subsection{Propagation effects}
So far, we have assumed that the electromagnetic radiation propagates through free space, resulting in an $r^{-2}$ fall-off. In reality, the radiation will be affected by complicated propagational effects. Because of the irregular surface of the Earth (including urban areas) and the absence of a direct line of sight between transmitter and receiver, the propagation mode might be indirect. Multiple indirect paths might be formed by reflection, refraction or diffraction of the electromagnetic wave.

A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}. The original model is based on urban areas, but corrections are given by \citeauthor{hata-propagation-loss} for sub-urban areas with a lower population density and for open areas. \citeauthor{hata-propagation-loss} gives the following analytical estimate for $L_p$, the electromagnetic propagation loss over land:
\begin{align} 
\notag                L_p & = & 69.55 + 26.16 \log_{10} f_c - 13.82 \log_{10} h_b - \\
\label{eq:hata-model} & &  a(h_m) + (44.9 - 6.55 \log_{10} h_b) \log_{10} r, 
\end{align}
where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in meters; $h_m$ the height of the receiving antenna in meters; $r$ the distance between the antennas in meters; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the urban density. \citeauthor{hata-propagation-loss} found this model to be representative for frequencies $f_c \sim$~150--1500~MHz, with transmitter heights $h_b \sim$~30--200~m, receiver heights $h_m \sim$~1--10~m and over distances $r\sim$~1--20~km.

@@

So far, these are the expected histograms for pure noise and pure RFI that propagates through free space. However, the measured distribution is a mixture of the two. A perfect RFI detector would separate the samples in two distributions; one that is not affected by RFI, and therefore contains noise only, and one that is the sum of RFI and noise. Real RFI detectors can separate these distributions to some extent, but due to false positives and false negatives, the histograms will get mixed nevertheless. Even more, we still need to take into account that the samples classified as RFI are also affected by noise. Although this effect is moderate, because 'most' RFI samples are of much higher amplitude than what is added because of the noise, we are interested in the low-level RFI samples, and the relative effect of noise on those samples is large.

@@

To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as analytic derivation is not trivial. The histogram can be numerically estimated by drawing complex real and imaginary samples from the two distributions and calculating and counting the amplitudes. A sample can be drawn from the RFI distribution by normalizing and inverting the cumulative frequency function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable. Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity. Therefore, to draw a sample by using the inversion method, this would require a uniformly distributed variable $\sim U(0,\infty)$. By assuming that there are no samples beyond some limiting distance $r_L$, then $F_{\textrm{amplitude}\le S}(S) = N-c \pi r_L^2$ for $S < Ig/r_L^2$. Consequently, by sampling $x_u$ from a uniform distribution with $0 \le x_u \le N-c \pi r_L^2$ and applying the inverse of Eq.~\eqref{eq:cumulative-rfi-distribution}, we will sample amplitudes $S$ with $S \ge Ig/r_L^2$, and $S\sim$ the power-law distribution.

By normalizing $x_u$, a single complex RFI contaminated sample $S_{\textrm{RFI}}$ can be sampled with
\begin{equation} \label{eq:drawing-rfi-distribution}
\textrm{real}(S_{\textrm{RFI}})\leftarrow \frac{Ig}{x_u r_L^2}\textrm{,\hspace{1mm} imag}(S_{\textrm{RFI}}) \leftarrow \frac{Ig}{y_u r_L^2},
\end{equation}
with $x_u, y_u \sim U(0, 1)$, and the corresponding formula for drawing a sample $S$ that is contaminated by both RFI and noise is
\begin{equation} \label{eq:drawing-combined-distribution}
\textrm{real}(S) \leftarrow x_n + \frac{Ig}{x_u r_L^2}
\textrm{,\hspace{1mm} imag}(S) \leftarrow y_n + \frac{Ig}{y_u r_L^2},
\end{equation}
with $x_n \sim N(\mu=0; \sigma=1)$. The final amplitude sample $S_A$ can be calculated with 
\begin{equation} \label{eq:drawing-amplitudes}
S_A \leftarrow |S| = \sqrt{\textrm{real}({S})^2+\textrm{imag}({S})^2}.
\end{equation}
An example of distribution curves for various settings of $Ig/r_L^2$ is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.

@@

One instrumental effect that is absorbed in $g$ is the frequency response of the instrument, i.e., the antenna response in combination with the band-pass of the analogue and digital filters. Because the data that are analysed in Sect.~\ref{sec:dist-results} have initially not been band-pass calibrated, the instrumental response is not uniform over frequency. We determined that the variation due to the band-pass is about one order of magnitude for the low-band antennas (LBA, 30.1--77.5 MHz) and about a factor of two for the high-band antennas (HBA, 115.0--163.3 MHz).

@@

%In Fig.~\ref{fig:plot-passband-histogram}, the histograms of the frequency response are given. To also demonstrate the effect of mixing distributions, histograms are added for samples that are drawn from a uniform, Gaussian and -2 power-law distribution as in Eq.~\eqref{eq:drawing-rfi-distribution} with $Ig/r_L^2=1$, that have been multiplied by the band-pass. The frequency responses were determined by measuring the standard deviation of the data over frequency after flagging the data for interference. As expected, it can still be determined that the intrinsic distribution is a -2 power law.

Another effect that is absorbed in $g$, is the beam of the instrument. At the point of writing, LOFAR beam models are still being developed and are not yet well parametrized. It is likely that most RFI sources are observed at the edges of the beam. Because most sources will be observed with similar gains (within one order of magnitude), it can be expected that the beam will have a limited effect on the histogram properties of an observation, and is comparable with the effect of the frequency response.

@@

\subsection{Estimating $\sigma$ and slope parameters}
The mode $\sigma$ of the Rayleigh distribution is estimated by finding the amplitude with the maximum occurrences, i.e., the amplitude corresponding to the peak of the histogram. The slope is estimated using linear regression over a visually selected interval.
%\begin{equation}
%R=\left[20\hat\sigma; 10^{\frac{1}{2}\left(\log_{10} 20\hat\sigma + \log_{10} S_\textrm{max}\right) } \right].
%\end{equation}
%In terms of the log-log histogram, the start of the interval corresponds with the point that is $1.3$ units to the right of the peak in the histogram, and the interval end corresponds with the point halfway between the start point and the maximum amplitude. This interval starts near the start of the RFI slope but past where the Rayleigh curve dominates. It also does not include the noisy data at the end of the curve. The curves are not always completely straight in this interval, but since we use the same stable way to determine the interval over which to calculate the slope, we can at least compare them.

One should note that fitting straight lines to the distribution curve in a log-log plot is not the most accurate way of estimating the exponent of a power-law distribution \citep{power-law-distribution}. However, because of our enormous sample size, which allows fitting the line over a large interval, the estimator will be sufficiently accurate for our purpose. Nevertheless, we will additionally calculate a maximum likelihood estimator for comparison. The maximum-likelihood estimator for the exponent in a power-law distribution is given by the Hill estimator \citep{hill-estimator, power-law-distribution}, defined as:
\begin{equation} \label{eq:hill}
 \alpha_H = 1 + N \left(\sum\limits_{i=1}^{N} \ln \frac{x_i}{x_\textrm{max}} \right)^{-1}.
\end{equation}
However, this estimator assumes the distribution is not cut-off at a high point. In our case, the distribution is cut-off at both sides, for example because of the limit of the analogue-digital converter (see \S\ref{sec:rfi-distribution-constraints}). Therefore, using this estimator will result in an estimate that is steeper (i.e., more negative) than the actual distribution. Moreover, the estimator requires an iteration over all data points, and because of the size of the data sets this is somewhat impractical. However, we can calculate Eq.~\eqref{eq:hill} by adding the bin centre values $N$ times to the calculation, where $N$ the bin frequency. Because our bin size is small and the distribution is well sampled, the resulting estimate will be close to the normal Hill estimator. This avoids the need to iterate over all data.

@@

Similar to the calculation of $S_U$ and by using $N_\textrm{total}$ as an upper limit to the total number of samples that are affected by the power-law distribution, one can estimate the lower limit $S_L$. For Fig.~\ref{fig:explanation-constraints} this means that the areas labelled $N_a$ and $N_\textrm{total}-N_\textrm{part}$ are equal. The area labelled $N_b$ is assumed to be zero for now, which assumes the power law has a sharp cut-off on the left side (e.g., because of the curvature of the Earth). Solving this equality, one finds that
\begin{equation} \label{eq:lower-limit-1}
S_L = \sqrt[\alpha+1]{\frac{\alpha+1}{\beta}\left(N_\textrm{part} - N_\textrm{total} \right) + S_1^{\alpha+1}}.
\end{equation}

@@

\section{Data description} \label{sec:dist-data}
We have analysed the distributions of two data sets. Both data sets are 24 h LOFAR RFI surveys and are extensively analysed in
\citet{lofar-radio-environment}. % TODO
In one set, the low-band antennas (LBA) were used and the frequency range 30.1--77.5~MHz was recorded, while in the other the high-band antennas (HBA) were used to record the frequency range 115.0--163.3~MHz. More stations were used in the LBA set. The specifications of the two sets are listed in Table~\ref{table:dist-data-specs}. The stations that have been used are geometrically spread over an area of about 80~km and 30~km in diameter at maximum for the LBA and HBA sets respectively.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=12cm]{img/histograms-raw-trimmed}
\caption{The histograms of the two data sets before band-pass correction and flux calibration.}
\label{fig:Histograms-raw}
\end{center}
\end{figure*}

@@

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/plot-lba-dist-per-frequency-trimmed}
\caption{Histograms for $5$ different $0.2$ MHz LBA sub-bands without band-pass correction and flux calibration. The continuous lines represent the data before RFI flagging. The dashed lines are the histograms of the samples that have been classified as RFI.}
\label{fig:plot-dist-per-frequency-LBA}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/plot-hba-dist-per-frequency-trimmed}
\caption{Histograms for $5$ different $0.2$ MHz HBA sub-bands without band-pass correction and flux calibration. The continuous lines represent the data before RFI flagging. The dashed lines are the histograms of the samples that have been classified as RFI.}
\label{fig:plot-dist-per-frequency-HBA}
\end{center}
\end{figure}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=13cm]{img/histograms-corrected-trimmed}
\caption{Observed LBA distribution after band-pass correction and flux calibration. $S_{l1}$ and $S_{l2}$ denote the limits of the distribution with a sharp lower cut-off (Eq.~\eqref{eq:lower-limit-1}) and uniform lower limit (Eq.~\eqref{eq:lower-limit-2}), $S_d$ is the average lower limit of RFI that is detected.}
\label{fig:histogram-passband-corrected}
\end{center}
\end{figure*}

To make sure that the antenna response does not influence the result of the slope, we have also analysed the curves after a simple band-pass calibration. This was performed by dividing each sub-band by its standard deviation after RFI excision. Because the standard deviation of the distribution might be affected by the RFI tail of the distribution, we compare the two histograms to make sure the power-law distribution is not significantly changed. The resulting histograms are shown in Fig.~\ref{fig:histogram-passband-corrected}. This procedure makes the bulge of the LBA histogram similar to the bulge of a Rayleigh curve and extends the power-law part. Nevertheless, it does not change the log-log slope of the power law in either histograms. This validates that the variable gain that is caused by the antenna response does not change the observed power law. Consequently, it can be expected that other stochastic effects, such as the intrinsic source strength and the beam gain due to a differing direction of arrival, will similarly not effect the power law. Because the band-pass corrected histograms provide a more representative analysis, we will use the corrected histograms for further analysis.
